---
title: "Class 7: Machine Learning 1"
author: "Kai Christman (A17522126)"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionality reduction**. The goal of **clustering** is to find groups within the data. 


## K-means clustering

The meain function in "base" R for K-means clustering is called `kmeans()`. 

Before we dive too deeply, let's make up some "simple" data that we can cluster and know if we are getting a good answer. To assist with this process we can use `rnorm()` function: 
```{r}
hist( rnorm(1000, mean = 3, sd = 1) )
```

```{r}
x <- c( rnorm(30, -3), rnorm(30, 3) )
x
```

The reverse function reverses the order of the vector, positives are now first and negatives are second. 
```{r}
z <- cbind(x = x, y = rev(x))
z

plot(z)
```

Now we can run `kmeans()` on this input `z` and see what the results look like!

```{r}
km <- kmeans(x = z, centers = 2)

km
```
```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What component of your result object details cluster assignment/membership?

```{r}
km$cluster
```

> Q. What component of your result obbject details cluster center?

```{r}
km$centers
```

> Q. Plot `z` colored by the kmeans cluster assignmnet and add cluster centers as blue points:

```{r}
plot(z, col = km$cluster)
points(km$centers, co = "blue", pch=15)

```

> Q. Run a K-means clustering and plot the results asking for 4 clusters (K=4):  

```{r}
km4 <- kmeans(x = z, centers = 4)

km4$cluster

plot(z, col = km4$cluster)
points(km4$centers, col = "orange", pch=15)

```


> **N.B.** You need to tell K-means the number of clusters (i.e. set of `centers = 2`)!!

One approach is to try different values for `centers` and pick the best...

```{r}
ans <- NULL

for(i in 1:10) {
  km <- kmeans(z, centers = i)
  ans <- c(ans, km$tot.withinss)
 
}

plot(ans, typ="o", 
     xlab = "Number of clusters", 
     ylab = "Total Sum of Squares")
```


## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`. This will work in bottom up fashion. 
Starts with each data point being in its own cluster, grouping some of them together until everything is in one cluster at the end. 

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`

```{r}
d <- dist(z)
hc <- hclust(d)
```

This is a bespoke `plot()` method for `hclust()` result objects. 
```{r}
plot(hc)
```

```{r}
plot(hc)
abline(h=8, col = "red")
```

Once we have our `hclust` object (our "tree" of cluster dendrogram") we can *"cut"* the tree to reveal the clustering pattern. 

```{r}
cutree(hc, h=8)
```

> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership) 

```{r}
groups <- cutree(hc, k = 2)
plot(z, col = groups)
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction method that is popular for revealing patterns in complex data sets. 
PCA takes multi-dimensional datasets and project them into useful ways, such as making figures. 
PCA aims to reduce dimensionality in an effective way, not all columns are necessary to evaluate the data.
In the most simplistic model, with just x and y dimensions, PCA fits a line between data, such as a regression. 

## Analysis of UK Food Data

Lets look at some data on the eating habbits of people from the UK and evalaute them to see if any patterns or trends that have some regions being distinct from others. 




##Data Import 

the dad is made available in CSV formate so we can use `read.csv()` to import into R:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
nrow(x)
```

## Tidy Data

Fix anything that went wrong with data import. 

## Previewing the First 6 rows of the data
```{r}
head(x)
dim(x)


rownames(x) <- x[,1]
x <- x[,-1]
head(x)
dim(x)
```
It is possible to set the row.names when reading the CSV file in, such as the code below:
```{r}
x <- read.csv(url, row.names=1)
head(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second approach as it allows the rows to be properly assigned at the start of data analysis when the CSV files are loaded. In addition, if you run the first example of code multiple times, it will continue to remove rows and labels, which could be detremental to your code. 


## Exploratory Analysis

Make some plots to help make sense of obvious trends ...

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```


> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```


```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```
**Key-point**: Even relatively small data sets can prove challenging to interpret Given that it is quite difficult to make sense of even this relatively small data set. Hopefully, we can clearly see that a powerful analytical method is absolutely necessary if we wish to observe trends and patterns in larger data sets.

## PCA
The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables" to be columns. 

This block of code will transpose "x" input objects
```{r}
pca <- prcomp( t(x) )
summary(pca)
```
The returned PCA object has components that we can use to make our main result figures: 

```{r}
attributes(pca)
```

The main result figure form this analysis is called a "PC Score plot" or "ordenation plot" or "PC Plot" or PC1 vs PC2 plot. How does your data lie along the new things that we have calculated? This is a reduced dimension space, reducing what is needed to describe the relationship between these countries, for exampls. 

This plot shows how samples (in this case countries) relate to each other along our new PC axis. (PC1 carries 67% of the variance, so it is more important as a representation of the data.)

```{r}
library(ggplot2)

ggplot(pca$x) +
  aes(PC1, PC2) + 
  geom_point()
```

```{r}
countrycols <- c("orange", "red", "blue", "darkgreen")

pca$x

ggplot(pca$x) +
  aes(PC1, PC2, label) + 
  geom_point( col = countrycols ) 

```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = row.names(pca$x) ) + 
  geom_point( col = countrycols ) +
  geom_text(size = 3, vjust=2, col=countrycols)
```

We can see that Northern Ireland is the most different as calcualted by PCA, but why?
Using `pca$rotation` we can observe all the calculations for each PC as calculated by PCA. 
```{r}
ggplot(pca$rotation) +
  aes(PC1, row.names(pca$rotation)) +
  geom_col()
```

The barchart of columns above shows that fresh potatoes and soft drinks contribute heavily to N. Irelands skewed results and high PCA variance. 

**IMPORTANT TAKEAWAYS** PC Score Plots, Loading Plots, and Scree plots are the most important takeaways from today's lab. A PC score plot shows how observations score within each principal component, with the axes usually representing the 2 most important components that help to clarify and explain trends in the data. A loading plot highlights all of the variables that are weighted by principal components and show what variables contribute heavily to the PC score. A scree plot describes variance, ultimately displaying each principal component and what the level of variance each new principal component will highlight. Most informaticians will use a Scree Plot to find the "elbow" flexion point to analyze which principal components are truly needed to fully analyze and interpret the data. 